Here is a beautifully structured, production-ready R Markdown vignette designed for your package documentation (`vignettes/BHiCeCT2-quickstart.Rmd`). It builds a cohesive, peer-reviewed story for an Application Note—taking the user smoothly from raw `.mcool` data extraction to multi-scale visualization and Bioconductor integration.

---

```markdown
---
title: "Multi-Scale Non-Contiguous Chromatin Partitioning with BHiCeCT2"
author: "Vipin Kumar"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Scale Non-Contiguous Chromatin Partitioning with BHiCeCT2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
        collapse = TRUE,
        comment = "#>",
        fig.width = 7,
        fig.height = 5,
        warning = FALSE,
        message = FALSE
)
```

## Introduction

Traditional HiC analysis restrict themselves to view HiC data at a single resolution. However, high-resolution Hi-C data shows that chromatin folds into a variety of configurations and looping networks spanning multiple scales best captured at correspondingly numerous resolutions.

`BHiCeCT2` addresses this biology via a **recursive spectral bisection algorithm** governed by an **iterative resolution-optimising framework**. If a localized chromatin domain looks uniform, the algorithm automatically zooms into a higher resolution to uncover nested topological details. If a boundary is found, it accepts the bipartition and continues partitioning at the current resolution.

This vignette demonstrates how to execute the core pipeline on an ultra-deep human HCT116 `.mcool` dataset, inspect diagnostic node statistics, generate publication-ready matrix overlays, and export clusters into the Bioconductor ecosystem.

---

## 1. Setup and Global Configurations

We begin by loading `BHiCeCT2` along with its core genomic and graph infrastructure dependencies. We disable scientific notation to keep large base-pair coordinate annotations human-readable.

```{r setup}
library(BHiCect2)
library(hictkR)
library(dplyr)
library(ggplot2)
library(GenomicRanges)

# Disable scientific notation for clean genomic positioning
options(scipen = 9999)
```

---

## 2. Multiscale Data Loading & Bisection Execution

We establish a connection to our hierarchical multi-resolution `.mcool` file using `hictkR`. We target **Chromosome 20**, a highly active, gene-dense archetype famous for complex looping topologies. We set a permissive decision gate threshold of `0.5`, making our significance baseline exactly equivalent to comparing observed boundary splits against the median noise of the bootstrapped null distribution. In essence, this criteria adapts the gap statistics to the specific task of Hi-C clustering.

```{r execution, eval = FALSE}
# Connect to the multi-resolution file pointer
mcool_path <- "/home/vipink/Documents/BHiCeCT2/data/HCT116/4DNFIP8RKGDG.mcool"
current_MresFile <- hictkR::MultiResFile(mcool_path)

# Execute the core recursive bisection pipeline
res_obj <- waterfall_bhicect(
        file = current_MresFile,
        chrom = "chr20",
        threshold = 0.5
)

# Persist output for downstream reporting consistency
saveRDS(res_obj, file = "./data/chr20_res_obj.rds")
```

```{r load_cache, include = FALSE}
# Internal helper simulating vignette rendering from cached tracking files
# res_obj <- readRDS("chr20_res_obj.rds")
# summary_tbl <- res_obj$nodes
```

---

## 3. Extracting and Reviewing the Tree Workspace

The result object contains the Nodes and edges of the clustering tree produced by BHiCect2 as two dedicated dataframes. We can focus on the Ndes or cluster summary table (`summary_tbl`), which includes tree depth, resolution level, empirical $p$-values (`perf`), and the raw list of genomic bin start coordinates (`ids`).

```{r summary_inspect}
# Load the pre-computed tree architecture
data("res_obj", package = "BHiCect2")
# Access the data frame from the loaded object
summary_tbl <- res_obj$nodes
# Preview the multi-tier data structural layout
head(summary_tbl |> dplyr::select(id, res_level, depth, perf, obs))
```

---

## 4. Visualizing Hierarchical Cluster Overlays

Using our package's native `plot_cluster_heatmap()` engine, we project the found clusters into explicit interaction matrices. The colormap of the heatmap reflects the tree depth of the considered cluster.

Nested TAD-like features and loops show through via alpha transparency blending, cleanly exposing how our multi-resolution description captures a variety of nested interaction patterns.

```{r plot_heatmap, fig.align='center', eval=FALSE}
# Generate the dynamic matrix block visualization
plot_cluster_heatmap(
        summary_tbl = summary_tbl,
        current_MresFile = current_MresFile
) +
        labs(title = "BHiCeCT2 Hierarchical Sub-TAD Map", subtitle = "Chromosome 20")
```

---

## 5. Statistical Diagnostics for Node Separability

To evaluate if an accepted split is physically sound, `BHiCeCT2` evaluates the observed separability score against a Student-$t$ distribution fit to a bootstrapped null background.

Because our separability score metric is bounded at $1.0$, our diagnostic function `plot_node_density_boot()` explicitly handles metric censoring. It captures leaking probability mass beyond the mathematical ceiling and renders it as a discrete **Dirac spike** right at $1.0$.

```{r plot_diagnostic, fig.align='center'}
# Select a prominent deep tree node to view validation diagnostics
target_node <- "D4_R5_29_30000000_44000000"

plot_node_density_boot(
        node_id = target_node,
        summary_tbl = summary_tbl,
        fill_tail = TRUE
)
```

---

## 6. Bioconductor Ecosystem Integration via `GRangesList`

To seamlessly connect our non-contiguous structural networks to downstream genomic workflows (such as intersecting with CTCF ChIP-seq peaks or checking enrichment profiles via `diffHic`), we parse the summary data frame into a structured `GRangesList` object.

The `as_granges_list()` converter automatically determines the appropriate scale-aware break margins based on each node's active resolution footprint.

```{r export_ranges, eval = FALSE}
# Convert non-contiguous cluster coordinate lists to a Bioconductor GRangesList
cluster_GRanges <- as_granges_list(
        summary_tbl = summary_tbl,
        current_MresFile = current_MresFile,
        seqname = "chr20"
)

# Inspect the structured S4 Bioconductor output
cluster_GRanges
```

We can now cleanly calculate the exact multi-fragment genomic footprint for any specific non-contiguous cluster hub:

```{r downstream_example, eval =FALSE}
# Quantify the total base-pair footprint of the target cluster hub across gaps
sum(GenomicRanges::width(cluster_GRanges[[target_node]]))
```

---

## Session Information

For transparency and reproducibility reporting standards:

```{r session_info}
sessionInfo()
```

```

```
