Here is a beautifully structured, production-ready R Markdown vignette designed for your package documentation (`vignettes/BHiCeCT2-quickstart.Rmd`). It builds a cohesive, peer-reviewed story for an Application Note—taking the user smoothly from raw `.mcool` data extraction to multi-scale visualization and Bioconductor integration.

---

```markdown
---
title: "Multi-Scale Non-Contiguous Chromatin Partitioning with BHiCeCT2"
author: "Vipin Kumar"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Scale Non-Contiguous Chromatin Partitioning with BHiCeCT2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
        collapse = TRUE,
        comment = "#>",
        fig.width = 7,
        fig.height = 5,
        warning = FALSE,
        message = FALSE
)
```

## Introduction

Traditional Topologically Associating Domain (TAD) callers treat chromatin architecture as a linear sequence of contiguous physical blocks. However, high-resolution Hi-C data (such as $1\text{ kb}$ maps) shows that chromatin folds into non-contiguous structural hubs and looping networks separated by un-interacting gaps.

`BHiCeCT2` addresses this biology via a **recursive spectral bisection algorithm** governed by an **inverse resolution-shifting framework**. If a localized chromatin domain looks uniform (the bootstrapped separability metric fails to reject the null distribution), the algorithm automatically zooms into a higher resolution tier to uncover nested topological details. If a boundary is found, it accepts the bipartition and continues partitioning.

This vignette demonstrates how to execute the core pipeline on an ultra-deep human HCT116 `.mcool` dataset, inspect diagnostic node statistics, generate publication-ready matrix overlays, and export clusters into the Bioconductor ecosystem.

---

## 1. Setup and Global Configurations

We begin by loading `BHiCeCT2` along with its core genomic and graph infrastructure dependencies. We disable scientific notation to keep large base-pair coordinate annotations human-readable.

```{r setup}
library(BHiCeCT2)
library(hictkR)
library(dplyr)
library(ggplot2)
library(GenomicRanges)

# Disable scientific notation for clean genomic positioning
options(scipen = 9999)
```

---

## 2. Multiscale Data Loading & Bisection Execution

We establish a connection to our hierarchical multi-resolution `.mcool` file using `hictkR`. We target **Chromosome 20**, a highly active, gene-dense archetype famous for complex looping topologies. We set a permissive decision gate threshold of `0.5`, making our significance baseline exactly equivalent to comparing observed boundary splits against the average expected noise of the bootstrapped null distribution.

```{r execution, eval = FALSE}
# Connect to the multi-resolution file pointer
mcool_path <- "/home/vipink/Documents/BHiCeCT2/data/HCT116/4DNFIP8RKGDG.mcool"
current_MresFile <- hictkR::MultiResFile(mcool_path)

# Execute the core recursive bisection pipeline
res_obj <- waterfall_bhicect(
        file = current_MresFile,
        chrom = "chr20",
        threshold = 0.5
)

# Persist output for downstream reporting consistency
saveRDS(res_obj, file = "chr20_res_obj.rds")
```

```{r load_cache, include = FALSE}
# Internal helper simulating vignette rendering from cached tracking files
# res_obj <- readRDS("chr20_res_obj.rds")
# summary_tbl <- res_obj$nodes
```

---

## 3. Extracting and Reviewing the Tree Workspace

The execution object contains structural metadata across all recursive bisection actions. We isolate the structured node execution summary table (`summary_tbl`), which tracks tree paths, active resolution index levels, empirical $p$-values (`perf`), and the raw list of genomic bin start coordinates (`ids`).

```{r summary_inspect}
# Load the pre-computed tree architecture
res_obj <- readRDS("chr20_res_obj.rds")
summary_tbl <- res_obj$nodes

# Preview the multi-tier data structural layout
head(summary_tbl |> dplyr::select(id, res_level, depth, perf, obs))
```

---

## 4. Visualizing Hierarchical Cluster Overlays

Using our package's native `plot_cluster_heatmap()` engine, we flatten the tree summaries into explicit multi-scale matrices. The function maps continuous fragments within each cluster relative to its resolution tier, packing them into clear geometrical footprints.

Nested sub-TAD features show through via alpha transparency blending, cleanly exposing how our resolution-shifting loop breaks uniform domains into crisp micro-domains.

```{r plot_heatmap, fig.align='center'}
# Generate the dynamic matrix block visualization
plot_cluster_heatmap(
        summary_tbl = summary_tbl,
        current_MresFile = current_MresFile
) +
        labs(title = "BHiCeCT2 Hierarchical Sub-TAD Map", subtitle = "Chromosome 20 Workspace")
```

---

## 5. Statistical Diagnostics for Node Separability

To evaluate if an accepted split is physically sound, `BHiCeCT2` evaluates the observed separability score against a Student-$t$ distribution fit to a bootstrapped null background.

Because our separability score metric is bounded at $1.0$, our diagnostic function `plot_node_density_boot()` explicitly handles metric censoring. It captures leaking probability mass beyond the mathematical ceiling and renders it as a discrete **Dirac spike** right at $1.0$.

```{r plot_diagnostic, fig.align='center'}
# Select a prominent deep tree node to view validation diagnostics
target_node <- "D4_R5_29_30000000_44000000"

plot_node_density_boot(
        node_id = target_node,
        summary_tbl = summary_tbl,
        fill_tail = TRUE
)
```

---

## 6. Bioconductor Ecosystem Integration via `GRangesList`

To seamlessly connect our non-contiguous structural networks to downstream genomic workflows (such as intersecting with CTCF ChIP-seq peaks or checking enrichment profiles via `diffHic`), we parse the summary data frame into a structured `GRangesList` object.

The `as_granges_list()` converter automatically determines the appropriate scale-aware break margins based on each node's active resolution footprint.

```{r export_ranges}
# Convert non-contiguous cluster coordinate lists to a Bioconductor GRangesList
cluster_GRanges <- as_granges_list(
        summary_tbl = summary_tbl,
        current_MresFile = current_MresFile,
        seqname = "chr20"
)

# Inspect the structured S4 Bioconductor output
cluster_GRanges
```

We can now cleanly calculate the exact multi-fragment genomic footprint for any specific non-contiguous cluster hub:

```{r downstream_example}
# Quantify the total base-pair footprint of the target cluster hub across gaps
sum(GenomicRanges::width(cluster_GRanges[[target_node]]))
```

---

## Session Information

For transparency and reproducibility reporting standards:

```{r session_info}
sessionInfo()
```

```

```
